Chemoinformatics: The Computer Science of Chemistry

### 2011-06-21 paper.tex
some fixes here and there
Added some text to highight the "stort" were trying present

Novel technologies in the life sciences churn out information at an ever
increasing rate – public data stores such as at the European Bioinformatics
Institute (EBI) contain on the order of 10 Petabytes of biological information,
a long way from the few kilobytes of the first sequenced genome in 1977.
However, while biological information has been available publicly for decades,
the same has not been true for chemical information until very recently. Only
2004 saw a large public compound repository, PubChem, entering the public
domain, followed more recently by other databases as discussed later in this
article. In the same vein, open source software and the publication of
algorithms have only recently become a main focus of the field \cite{faulon2010}.

@@

Both of these questions have relatively straightforward answers. We should care
about chemical information being made public since the drugs we take are the
result of intensive and costly research, and the more information is shared (a
process sometimes very hard to achieve in pharmaceutical companies) the more
easy it should be to develop novel treatments drugs. How chemical information
relates to computer science becomes clear when realizing the amount of data
available currently -- 74 million different molecules with annotations are
currently stored in public databases such as PubChem with a large amount of
annotations, and the design of data structures for chemicals, as well as
subsequent data mining of this information, are areas where expertise in
computer science is dearly needed. (For a recent, detailed introduction to the
field written for computer scientists see \cite{brown2009}). Cheminformatics
comprises different areas, which broadly speaking can be divided into three
fields: capturing data (using lab notebooks or potentially using formats such as
the Chemical Markup Language for publications); storing data (designing database
schemas, devising ontologies) and mining data (such as for bioactivity
prediction, which might involve algorithms based on graphs). However, what is a
major difference to the bioinformatics field is the type of information being
analyzed: While bioinformatics often deals with sequences, the domain of
cheminformatics is chemical structures. In case of the former, information can
very frequently be represented as one-dimensional strings, which are relatively
easy to handle computationally. On the other hand, chemical structures may
possess rings, branches as well as multiple valid representations of even the
same structure, potentially giving rise to ambiguities. Hence, chemical
structure is often thought of as being more difficult to standardize than
information from other domains. As an illustration, different representations of
the same chemical structure are shown in Figure 1. The top left hand corner
shows a structure a chemist would draw it -- every 'corner' representing a carbon
atom (which can also be represented as a C), every O denoting oxygen, every N
nitrogen, every H hydrogen and so on. This is the standard format chemists use
when exchanging information with their peers. However, this is not directly
amenable to data mining -- different atoms might be collapsed into one node,
sets of atoms might have the same relevant properties assigned. Hence, the top
right hand corner of Figure 1 shows an annotated graph representation, taking
into account that atoms, though being of different elemental atom type, can have
similar properties -- e.g. both oxygen and nitrogen carry an 'On' label in this
graph.

@@

\section{Bridging Cheminformatics \& Computer Science}
\label{sec:bridg-chem-}

\subsection{Databases \& Ontologies}
\label{sec:datab--ontol}

@@

@@

>>>>

Novel technologies in the life sciences churn out information at an ever
increasing rate -- public data stores such as at the European Bioinformatics
Institute (EBI) contain on the order of 10 PB of biological information.
However, while biological information has been available publicly for decades,
the same has not been true for chemical information until very recently. It was
not until 2004 that a large public small molecule structure repository (Pubchem)
became freely available. This was followed by other databases as discussed later
in this article. In the same vein, open source software and the publication of
algorithms have only recently become a main focus of the field
\cite{faulon2010}.

@@

Both of these questions have relatively straightforward answers. We should care
about chemical information being made public since the drugs we take are the
result of intensive and costly research, and the more information is shared (a
process sometimes very hard to achieve in pharmaceutical companies) the more
easy it should be to develop novel treatments drugs. How chemical information
relates to computer science becomes clear when realizing the amount of data
available currently -- 74 million different molecules with annotations are
currently stored in public databases such as PubChem with a large amount of
annotations, and the design of data structures for chemicals, as well as
subsequent data mining of this information, are areas where expertise in
computer science is dearly needed. (For a recent, detailed introduction to the
field written for computer scientists see \cite{brown2009}).

Cheminformatics comprises different areas, which broadly speaking can be divided
into three fields: capturing data (using lab notebooks or potentially using
formats such as the Chemical Markup Language for publications); storing data
(designing database schemas, devising ontologies) and mining data (such as for
bioactivity prediction, which might involve algorithms based on graphs).
However, what is a major difference to the bioinformatics field is the type of
information being analyzed: While bioinformatics often deals with sequences, the
domain of cheminformatics is chemical structures. In case of the former,
information can very frequently be represented as one-dimensional strings, which
are relatively easy to handle computationally. On the other hand, chemical
structures may possess rings, branches as well as multiple valid representations
of even the same structure, potentially giving rise to ambiguities. Hence,
chemical structure is often thought of as being more difficult to standardize
than information from other domains. As an illustration, different
representations of the same chemical structure are shown in Figure 1. The top
left hand corner shows a structure a chemist would draw it -- every 'corner'
representing a carbon atom (which can also be represented as a C), every O
denoting oxygen, every N nitrogen, every H hydrogen and so on. This is the
standard format chemists use when exchanging information with their peers.
However, this is not directly amenable to data mining -- different atoms might
be collapsed into one node, sets of atoms might have the same relevant
properties assigned. Hence, the top right hand corner of Figure 1 shows an
annotated graph representation, taking into account that atoms, though being of
different elemental atom type, can have similar properties -- e.g. both oxygen
and nitrogen carry an 'On' label in this graph.

@@

\section{Bridging Cheminformatics \& Computer Science}

The motivation for this article is to highlight a variety of topics in
cheminformatics and present the computational and algorithmic problems that are
addressed i the field. The hope is that we will be able to highlight areas where modern CS
research could contribute to these problems.

While cheminformatics has a broad range of applicability, ranging from
agrochemical research and drug discovery to the design of novel materials, we
have chosen to use drug discovery as the context of this article. The remainder
of this article is structured around the concept of ``risk minimization'' in
drug discovery -- how does one minimize the chances of a small molecule failing
as a drug candidate and how do cheminformatics methods play a key role in this
efforts.

\subsection{Databases \& Ontologies}

@@

See e.g. role for the EBI and genome campus,
https://registration.hinxton.wellcome.ac.uk/display_info.asp?id=246 See also
impact of open chemical data, ChEMBL to future, e.g. ELIXIR
http://www.elixir-europe.org/page.php

@@

\section{Conclusions}
\label{sec:conclusions}
